---
title: "Updating a DuckDB Database with Missing Files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Updating a DuckDB Database with Missing Files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # long-running, network + S3 access; run interactively
)
```

## Overview

The companion tutorial, *"Identifying New Returns to Process,"* showed how to find the filings
that are **available** but not yet **processed** into a year's database. This
tutorial takes the next step: it processes those missing filings and folds them
into a fresh, up-to-date copy of the database.

The whole update is a single call:

```r
update_db(year, index, path)
```

Under the hood, `update_db()` orchestrates three stages:

1. **`find_missing_urls()`** — diff the index against the remote database to get
   the URLs that still need processing (see the previous tutorial).
2. **`build_database(..., is_update = TRUE)`** — flatten just those missing XML
   files into a small *temporary* DuckDB.
3. **`merge_databases()`** — combine the existing remote database with the
   temporary one, aligning schemas, into a new local `EFILE<YEAR>.duckdb`.

You can run the one-shot version or drive each stage by hand for more control.

## Prerequisites

```{r prereqs}
library(ef2)

YEAR <- 2023
PATH <- "C:/Users/username/DATA/DUCKDB_2025"   # local output location

# The index of AVAILABLE filings, pulled from the Data Commons and
# subset to the tax year of interest.
index   <- get_current_index_full()             # or get_current_index_batch()
index_y <- index[index$TaxYear == YEAR, ]
```

You need:

- an **index** of available filings from `get_current_index_full()` (or
  `get_current_index_batch()`), subset to the year via its `TaxYear` column,
- network access to the GT data lake (raw XML) and the NCCS S3 bucket
  (the existing `EFILE<YEAR>.duckdb`), and
- enough local disk at `path` to hold the merged database.

## The one-shot update

```{r update-oneshot}
out_path <- update_db(year = YEAR, index = index_y, path = PATH)
out_path
```

Expected console output:

```
🚀 Updating database for TaxYear 2023
🔎 Checking for missing URLs in TaxYear 2023
1289 missing URLs detected.
Building temporary database with 1289 missing files

📦 Creating new batch files for 2023
🧮 Processing 52 batches across 4 workers for 2023...
🔧 Merging databases for year 2023
Appended KEYS from temporary DB (schema aligned).
Appended FLATXML from temporary DB (schema aligned).
Appended ATTRIBUTES from temporary DB (schema aligned).
✅ Merged DB written to: C:/Users/username/DATA/DUCKDB_2025/EFILE2023.duckdb
📜 Logfile saved at: merge_log_2023.txt
🎯 Update complete for TaxYear 2023
```

If the database is already current, `update_db()` short-circuits:

```
No missing files found. Database is up to date.
```

and returns `NULL` without writing anything.

The result is a **new local** database at
`<path>/EFILE<YEAR>.duckdb` that contains everything from the published remote
database **plus** the newly processed filings.

## What each stage does

### 1. Find the gap

Identical to the previous tutorial — this is the diff that decides whether there
is any work to do:

```{r stage-find}
missing_urls <- find_missing_urls(year = YEAR, index = index_y)
length(missing_urls)
```

If `length(missing_urls) == 0`, stop here.

### 2. Build a temporary database

`build_database()` flattens the missing XML into a temporary DuckDB. The
`is_update = TRUE` flag tells it this is an incremental build rather than a
full-year rebuild. It batches the URLs (default `group.size = 25`) and processes
them across up to four parallel workers.

```{r stage-build}
temp_db_path <- build_database(
  year      = YEAR,
  urls      = missing_urls,
  path      = PATH,
  is_update = TRUE
)
```

Only the missing filings are processed, so this is fast compared with rebuilding
a full tax year.

### 3. Merge into the final database

`merge_databases()` attaches both the remote source database (read-only) and the
temporary database, then appends the new rows into the three core tables —
`KEYS`, `FLATXML`, and `ATTRIBUTES` — reconciling any column differences so the
schemas line up.

```{r stage-merge}
output_path <- file.path(PATH, paste0("EFILE", YEAR, ".duckdb"))

merge_databases(
  year         = YEAR,
  missing_urls = missing_urls,
  temp_db_path = temp_db_path,
  output_path  = output_path
)
```

Running the three stages by hand is equivalent to a single `update_db()` call —
use it when you want to inspect the temporary database before merging, or re-run
just the merge.

## Reading the merge log

Every merge appends a timestamped record to `merge_log_<YEAR>.txt`, including
per-table before/after row counts and durations:

```
=== Merge Log for TaxYear 2023 ===
Start Time: 2026-07-09 10:22:14
Missing URLs: 1289

KEYS       | 2026-07-09 10:22:41 | 1204331 → 1205620 rows | Duration: 3.11 sec
FLATXML    | 2026-07-09 10:24:02 | 88231145 → 88402910 rows | Duration: 79.4 sec
ATTRIBUTES | 2026-07-09 10:24:10 | 1204331 → 1205620 rows | Duration: 7.88 sec
Total Duration: 121.4 sec
Merge complete.
```

The `KEYS` row count going from 1,204,331 → 1,205,620 (a jump of 1,289 — exactly
the number of missing URLs) confirms every missing filing landed.

## Step 4 — Verify the update

Two quick checks confirm the merged database is complete and self-consistent.

**Re-run the diff.** After the update, the gap against the same index should be
zero. Point a fresh read at your *local* file (or re-publish and re-check the
remote):

```{r verify-diff}
inspect_ddb(output_path, table = "KEYS", n = 5)   # peek at the merged KEYS table
```

**Confirm the row count.** The number of unique returns should now match the
index total for the year:

```{r verify-count}
con <- DBI::dbConnect(duckdb::duckdb(), dbdir = output_path)
DBI::dbGetQuery(con, "SELECT COUNT(*) AS n FROM KEYS;")
DBI::dbDisconnect(con, shutdown = TRUE)
```

## Updating several years

Because `update_db()` no-ops when a year is already current, it's safe to loop
over an entire archive — only years with missing files do any real work:

```{r update-all}
years <- 2019:2023
index <- get_current_index_full()   # pull the full multi-year index once

for (y in years) {
  idx <- index[index$TaxYear == y, ]
  update_db(year = y, index = idx, path = PATH)
}
```

## What you learned

- `update_db(year, index, path)` runs the full missing-files workflow end to end.
- It finds the gap, builds **only** the missing filings, and merges them with the
  existing database into a new local `EFILE<YEAR>.duckdb`.
- Every run leaves a `merge_log_<YEAR>.txt` audit trail with per-table row deltas.
- The update is **idempotent**: re-running on a current database changes nothing.
